For submission to the Journal of Integrated Pest Management

### 2021-03-16 ofv_fife_2021.Rmd
Accepted almost all of the changes the authors suggested: clarified plant taxonomy, made it clear that there were more than one mite species recovered from plants. Removed first part of section on control, which will require a rewrite. Other minor edits to formatting, added a new reference Dr. Carillo suggested

### First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) virus, *Orchid fleck dichorhavirus* (Mononegavirales: Rhabdoviridae) infecting ornamental Nolinoideae in the Florida panhandle

@@

### Abstract
Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vectors of dichorhaviruses and transmit *Orchid fleck dichorhavirus* (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects 50+ spp. of Orchidaceae, Nolinoideae, and *Citrus*.
During June 2020, we observed chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) in the landscape of Leon County, Florida. We tested plant samples for a wide range of pathogens and ultimately identified the pathogen as *Orchid fleck dichorhavirus*. We took further samples from other *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagales: Nolinoideae) in Leon and Alachua county. All symptomatic plants tested positive for OFV. We recovered mites from OFV-infected plants, which belonged to *Brevipalpus californicus* sensu lato.
Florida has various mite species of *Brevipalpus* and a diverse array of susceptible plant species in the landscape, both native and introduced. We suggest that OFV already has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for many plants of economic importance for Florida.

### Keywords:
false spider mite, flat mite, *Brevipalpus*-transmitted viruses, *Liriope*, Ruscaceae, Asparagaceae, orchid, Orchidaceae, pests, ornamental plants

@@

OFV naturally infects 50+ spp of Orchidaceae [@Kitajima2010; @Peng2013], some Nolinoideae [@Mei2016, @Dietzgen2018b], and *Citrus* where it causes citrus leprosis-like symptoms [@BeltranBeltran2020; @GarciaEscamilla2018 @Roy2020]. Mechanical transmission of OFV is possible under lab conditions to various *Chenopodiaceae*, *Aizoaceae*, *Fabaceae*, and *Solanaceae* [@Chang1976; @Kondo2003; @Peng2013].

During June 2020, chlorotic ringspot symptoms were observed on the liriopogon Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in the landscape of Leon County, Florida (`r citef("fig_1")`). Liropogons belong to the plant family Asparagaceae, subfamily Nolinoideae, which includes various monocotyledonous lilliod plants [@Chase2009]. Liriopogons are native to southeastern Asia, encompassing the closely-related genera *Liriope* and *Ophiopogon*, which are characterized by evergreen, long grass-like leaves, and erect flowers which produce black, purple or blue, berry-like fruits, depending on the cultivar [@Lattier2014; @Fantz2015]. These characteristics have given rise to various common names, including: monkey grass, Aztec grass, lilyturf, mondo grass, etc. (*`r citet("tab_1")`*). The genus *Liriope* contains eight species, four of which are commonly cultivated, while the genus *Ophiopogon* consists of about 65 species, with only a handful are available for horticulture [@Lattier2014; @Nesom2010]. Liripogons are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

Plant leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. The plant samples were tested for a wide range of common plant pathogens, but the results were inconclusive. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The presence of *Orchid fleck dichorhavirus* was confirmed via generic one step conventional RT-PCR with R2-Dicho-GF and R2-Dicho-GR primers, amplifying ~800 nt of L-gene (RNA2) amplicon from a leaf sample of infected *Liriope* [@Kubo2009b]. RT-qPCR confirmed OFV and the RT-PCR product was sequenced using Sanger sequencing and High Throughput Sequencing [@Kondo2006]. Sanger sequencing of RT-PCR amplicons shared 98 \% nucleotide identity with OFV strains which infect orchids (GenBank Accession numbers: AB244418 and MK522807) [@Kondo2006; @Kondo2014; @Cook2019]. Further samples were taken from various symptomatic cultivars of *Liriope* spp., *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagales: Nolinoideae) during subsequent visits to the initial site of collection as well as other locations in Leon county. Samples were tested via RT-PCR at the NFREC (*`r citet("tab_1")`*) to confirm the presence of OFV.

Further surveys of Nolinoideae in Florida have revealed more sites with symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, we collected mites from symptomatic plants in Leon county which were observed with phase contrast microscopy. We encountered both eriophyoid mites and flat mites which were identified as *Brevipalpus californicus* (Banks) sensu lato, an identity which was confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. *Brevipalpus* mites have been previously associated with OFV [@BeltranBeltran2020; @GarciaEscamilla2018; @Dietzgen2018b] and similar diseases [@Kitajima2010] and are known to feed on a large variety of economically-important plants [@Childers2003a; @Akyazi2017]. Unfortunately, the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as low-temperature scanning electron microscopy (LT-SEM) [@Skoracka2015; @Leon2010]. With that in mind, we sent additional samples of the mites to the USDA-ARS in Beltsville to observe the mites with LT-SEM techniques, which agreed with both prior identifications of *B. californicus* s.l. (`r citef("fig_2")`).

The earliest mention of a virus which may have been OFV is @Ko1985, who detected nuclear inclusions caused by an undescribed bacilliform rhabdovirus. @Ko1985 described the spoke-wheel configurations typically associated with OFV [@Chang1976], in *Brassia* orchids, but unfortunately made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis. In addition, @Bratsch2015 detected *Brevipalpus* mites and their exoskeletons associated with OFV-infected plants. It appears that @Bratsch2015 did not make a conclusive species identification, but did cite @Kondo2003, suggesting that they suspected *B. californicus* as their vector.

OFV has been reported in other Nolinoideae in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour. [@Mei2016]. Our plants are thought to belong to different species of *Liriope* which are not *L. spicata* and we are not aware of any record which reports OFV infection in *Ophiopogon* plants. Unfortunately, liriopogons species are very similar in appearance and growth habit, with few useful characters used for their classification [@Fantz2008a]. Furthermore, the horticultural industry has created a diverse array of cultivars of these plants, which are often mislabeled [@Fantz2008a]. Aside from the taxonomic confusion created by humans, natural hybrids between *Ophiopogon* and *Liriope* have created a natural source of error for reconstructing phylogenies [@Zhou2009]. Together, these factors make it difficult to differentiate and identify plants in the landscape by visual inspection alone. These obfuscations of species identity may be accounted for in the future via sequence comparisons of the OFV-infected plants, but these comparisons are beyond the scope of our current report. Nonetheless, we are confident that ours is the first report of OFV infecting *Aspidistra elatior*; although [@Zheng2013] mentions the association of *B. californicus* with *A. elatior* they make no mention of OFV symptoms in this plant.

@@

Reducing the populations of the *Brevipalpus* mite vector may be able to reduce the spread of OFV by infective mites, but infected plants are not able to recover from the virus and must be rouged or removed. Rouging plants is a direct and efficient option to remove mite and virus populations quickly but increases in difficultly with the number of plants involved. Once infected plants have been removed, the objective will be to prevent re-infestation of the *Brevipalpus* vector via conventional pest management methods.

Most research on the integrated management of mites has been catered towards controlling various species of spider mites (Acari: Tetranychidae) [@Leeuwen2015], but many of those methods can be applied to *Brevipalpus* spp. as well:

Chemical control via synthetic acaricides is commonly used to control *Brevipalpus* spp. [@Leeuwen2015], but acaricides have some drawbacks. Many products control a broad class of arthropods, but often they do so indiscriminately, killing beneficial insects as well as pests [@Suckling2013]. Control strategies which rely heavily on chemical control have the potential to encourage pesticide resistance in mite populations and pesticide resistance has been reported in various *Brevipalpus* populations [@Alves2000; @Campos2002; @Omoto2000]. In addition, it is important to consider the interactions which may occur between different chemical applications or tank mixes [@Vechia2018]. Lastly, application costs of chemical controls can be large and increase the cost of production [@Rodrigues2000a], nevertheless chemical applications are valuable tools used to consider against *Brevipalpus*.

@@

Detecting OFV in Florida represents a concern for horticulturalists who grow susceptible plants of economic importance, including orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagales spp. which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including the famous Ghost Orchid, *Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe. Furthermore, OFV represents an obstacle to overcome for the burgeoning interest in cultivating *Vanilla* in southern Florida [@Chambers2019]. Lastly, some OFV isolates are known to be involved with nuclear types of citrus leprosis [@Roy2015a], which may be a cause for concern for the citrus industry. In fact, @Kitajima2011a found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013a]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@BeltranBeltran2020; @Kondo2003; @Knorr1968] and *B. obovatus* is a suspected vector as well [@Childers2003a]. All three mite species/complexes are present in Florida [@Childers2003a; @Akyazi2017] (`r citef("fig_4")`); therefore, it is critical to monitor the spread of Orchid Fleck Virus and its mite vector(s) to determine what risk this virus represents for plants in Florida and the surrounding regions.

>>>>

### First report of the *Brevipalpus*-transmitted (Trombidiformes: Tenuipalpidae) virus, *Orchid fleck dichorhavirus* (Mononegavirales: Rhabdoviridae) infecting ornamental Nolinoidaea in the Florida panhandle

@@

### Abstract
Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vectors of dichorhaviruses and transmit *Orchid fleck dichorhavirus* (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects 50+ spp. of Orchidaceae, Asparagaceae (Nolinoidaea), and Rutaceae (*Citrus*).
During June 2020, chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) were observed in a landscape in Leon County, Florida. Later in the year, the presence of *Orchid fleck dichorhavirus* was confirmed via generic one step conventional RT-PCR and Sanger sequencing. Amplicons had a 98\% identity with OFV strains which infect orchids. The identification was also confirmed with RT-qPCR and High Throughput Sequencing (HTS). Additional samples were taken from other *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) in Leon and Alachua county. All symptomatic plants tested positive for OFVs. Three mite species were recovered from OFV-infected plants: *Brevipalpus californicus* sensu lato, *B. obovatus* Donnadieu and *B. confusus* Banks. One of these species is presumably responsible for OFV transmission. Florida has various mite species of *Brevipalpus* and a diverse array of susceptible plant species in the landscape, both native and introduced. We suggest that OFV already has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for *Liriope* spp., *Ophiopogon* spp. and *Aspidistra elatior*, which are commonly used in landscaping in Florida.

### Keywords:
False spider mite, flat mite, *Brevipalpus*-transmitted viruses, *Liriope*, Ruscaceae, Asparagaceae, orchid, Orchidaceae, pests, ornamental plants, orchid fleck virus

@@

OFV naturally infects 50+ spp of Orchidaceae [@Kitajima2010; @Peng2013], some Asparagaceae (Nolinoidaea) [@Mei2016; @Dietzgen2018b], as well as Rutaceae: *Citrus* where it causes citrus leprosis-like symptoms [@BeltranBeltran2020; @GarciaEscamilla2018; @Roy2020]. Mechanical transmission of OFV is possible under lab conditions to various Chenopodiaceae, Aizoaceae, Fabaceae, and Solanaceae [@Chang1976; @Kondo2003; @Peng2013].

During June 2020, chlorotic ringspot symptoms were observed on the liriopogon Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in a landscape of Leon County, Florida (`r citef("fig_1")`). Liropogons belong to the plant family Asparagaceae, subfamily Nolinoidaea, which includes various monocotyledonous lilliod plants [@Chase2009]. Liriopogons are native to southeastern Asia, encompassing the closely-related genera *Liriope* and *Ophiopogon*, which are characterized by evergreen, long grass-like leaves, and erect flowers which produce black, purple or blue, berry-like fruits, depending on the cultivar [@Lattier2014; @Fantz2015]. These characteristics have given rise to various common names, including: monkey grass, Aztec grass, lilyturf, mondo grass, etc. (*`r citet("tab_1")`*). The genus *Liriope* contains eight species, four of which are commonly cultivated, while the genus *Ophiopogon* consists of about 65 species, with only a handful are available for horticulture [@Lattier2014; @Nesom2010]. Liripogons are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

Plant leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. The plant samples tested negative for begomovirus, potyvirus, tospovirus, as well as INSV, TMV and TSWV. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The presence of *Orchid fleck dichorhavirus* was confirmed via generic one step conventional RT-PCR with R2-Dicho-GF and R2-Dicho-GR primers, amplifying ~800 nt of L-gene (RNA2) amplicon from a leaf sample of infected *Liriope* [@Kubo2009b]. RT-qPCR confirmed OFV and the RT-PCR product was sequenced using Sanger sequencing and High Throughput Sequencing [@Kondo2006]. Sanger sequencing of RT-PCR amplicons shared 98\% nucleotide identity with OFV strains which infect orchids (GenBank Accession numbers: AB244418 and MK522807) [@Kondo2006; @Kondo2014; @Cook2019]. Further samples were taken from various symptomatic cultivars of *Liriope* spp., *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagaceae: Nolinoidaea) during subsequent visits to the initial site of collection as well as other locations in Leon county. Samples were tested via RT-PCR at the NFREC (`r citet("tab_1")`) to confirm the presence of OFV.

Further surveys of Nolinoidaea in Florida have revealed more sites with symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, we collected mites from symptomatic plants in Leon county which were observed with phase contrast microscopy. We encountered both eriophyoid mites and flat mites which were originally identified as *Brevipalpus californicus* (Banks) sensu lato, an identity which was confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. *Brevipalpus* mites have been previously associated with OFV [@BeltranBeltran2020; @GarciaEscamilla2018; @Dietzgen2018b] and similar diseases [@Kitajima2010] and are known to feed on a large variety of economically-important plants [@Childers2003a; @Akyazi2017]. Unfortunately, the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as low-temperature scanning electron microscopy (LT-SEM) [@Skoracka2015; @Leon2010]. With that in mind, we sent additional samples of the mites to the USDA-ARS in Beltsville to observe the mites with Cryo LT-SEM techniques, which agreed with both prior identifications of *B. californicus* s.l. (`r citef("fig_2")`) but also revealed the presence of two other species: *B. obovatus* and *B. confusus*.

The earliest mention of a virus which may have been OFV is @Ko1985, who detected nuclear inclusions caused by an undescribed bacilliform rhabdovirus. @Ko1985 described the spoke-wheel configurations typically associated with OFV [@Chang1976], in *Brassia* orchids, but unfortunately made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis. In addition, @Bratsch2015 detected *Brevipalpus* mites and their exoskeletons associated with OFV-infected plants. It appears that @Bratsch2015 did not make a conclusive species identification, but did cite @Kondo2003, suggesting a mite in the *B. californicus* group as the vector.

OFV has been reported in other Nolinoidaea in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour. [@Mei2016]. Our plants are thought to belong to different species of *Liriope* which are not *L. spicata* and we are not aware of any record other than this manuscript which reports OFV infection in *Ophiopogon* plants. Unfortunately, liriopogons species are very similar in appearance and growth habit, with few useful characters used for their classification [@Fantz2008a]. Furthermore, the horticultural industry has created a diverse array of cultivars of these plants, which are often mislabeled [@Fantz2008a]. Aside from the taxonomic confusion created by humans, natural hybrids between *Ophiopogon* and *Liriope* have created a natural source of error for reconstructing phylogenies [@Zhou2009]. Together, these factors make it difficult to differentiate and identify plants in the landscape by visual inspection alone. These obfuscations of species identity may be accounted for in the future via sequence comparisons of the OFV-infected plants, but these comparisons are beyond the scope of our current report. Nonetheless, we are confident that ours is the first report of OFV infecting *Aspidistra elatior*; although [@Zheng2013] mentions the association of *B. californicus* with *A. elatior* they make no mention of OFV symptoms in this plant.

@@

<!-- Reducing the populations of the *Brevipalpus* mite vector may be able to reduce the spread of OFV by infective mites, but infected plants are not able to recover from the virus and must be rouged or removed. Rouging plants is a direct and efficient option to remove mite and virus populations quickly but increases in difficultly with the number of plants involved. Once infected plants have been removed, the objective will be to prevent re-infestation of the *Brevipalpus* vector via conventional pest management methods. -->

<!-- Most research on the integrated management of mites has been catered towards controlling various species of spider mites (Acari: Tetranychidae) [@Leeuwen2015], but many of those methods can be applied to *Brevipalpus* spp. as well: -->
Chemical control via synthetic acaricides is commonly used to manage *Brevipalpus* spp. [@Leeuwen2015], but acaricides have some drawbacks. Many products control a broad class of arthropods, but often they do so indiscriminately, killing beneficial insects as well as pests [@Suckling2013]. Control strategies which rely heavily on chemical control have the potential to encourage pesticide resistance in mite populations and pesticide resistance has been reported in various *Brevipalpus* populations [@Alves2000; @Campos2002; @Omoto2000; @Rocha2021]. In addition, it is important to consider the interactions which may occur between different chemical applications or tank mixes [@Vechia2018]. Lastly, application costs of chemical controls can be large and increase the cost of production [@Rodrigues2000a], nevertheless chemical applications are valuable tools used to consider against *Brevipalpus*.

@@

Detecting OFV in Florida represents a concern for horticulturists who grow susceptible plants of economic importance, including orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagaceae spp. which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including the famous Ghost Orchid, *Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe. Furthermore, OFV represents an obstacle to overcome for the burgeoning interest in cultivating *Vanilla* in southern Florida [@Chambers2019]. Lastly, some OFV isolates are known to be involved with nuclear types of citrus leprosis [@Roy2015a], which may be a cause for concern for the citrus industry. In fact, @Kitajima2011a found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013a]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@BeltranBeltran2020; @Kondo2003; @Knorr1968] and *B. obovatus* is a suspected vector as well [@Childers2003a]. All three mite species/complexes are present in Florida [@Childers2003a; @Akyazi2017] (`r citef("fig_4")`); therefore, it is critical to identify the vector of OFVs in Florida and monitor its spread to determine what risk this virus represents for plants in Florida and the surrounding regions.
